AAAI Press Formatting Instructions \\for Authors Using \LaTeX{} --- A Guide

### 2021-09-16 publications/iaai21/dora_outlier_detection_iaai.tex
Add description of PAE. #42

\subparagraph{Likelihood.}
The negative sampling algorithm is implemented by converting the unsupervised 
outlier ranking problem into a semi-supervised 
problem~\citep{sipple:neg-sampling20}. Negative (anomalous) 
examples are created by sampling from an expanded space defined by the minimum 
and maximum values of each dimension of the positive (normal) examples. The 
negative and positive examples are then used to train a random forest 
classifier. We use the posterior probabilities of the random forest classifier 
as outlier scores, which means that the observations with higher posterior 
probabilities are more likely to be outliers. \todo{Describe PAE: Hannah or Bryce}

@@

\subsubsection{Satellite time series for ground-truth observations.}
Many Earth science applications using satellite Earth observation (EO) data
 require ground-truth observations for understanding and modeling ground-
 identified objects in the satellite observations. These ground observations
 also serve as labels that are paired with satellite data inputs for machine 
 learning models. For example, a model trained to classify crop types in
 satellite observations requires geo-referenced labels of crop type annotated
 in the field~\cite{tseng2021learning}. A widespread challenge of using ground-
 annotated datasets is that there are often points with erroneous location or 
 label information (e.g., due to GPS location error or human entry error) that
 need to be cleaned before the dataset can be used for machine learning or
 other downstream use cases. Automatically detecting these outliers could save 
 substantial time required for cleaning datasets and improve the performance of
 downstream analyses that rely on high-quality datasets. 
 %
 We used a dataset of ground
 observations of fall army worm (FAW) in maize crops collected by the UN Food
 and Agriculture Organization (FAO). This dataset includes 6,757 samples
 with location (latitude/longitude) metadata primarily in Africa and Southeast
 Asia. Most locations coincide with crop
 fields but there are many outliers that coincide with other land cover types 
 such as water, buildings, or forests. We constructed an evaluation set 
 consisting of all samples in Kenya ($n=113$) and manually annotated whether 
 each sample was considered an inlier (i.e., located in a crop field) or not 
 using high-resolution Planet satellite images in Collect Earth 
 Online~\cite{planetdata} ($n_{inlier}=76$, $n_{outlier}=37$). We used Google
 Earth Engine to export the Sentinel-1 synthetic aperture radar (SAR) monthly
 time series for each sample location, using data from year each sample was 
 collected. We used SAR data because it is sensitive to ground texture and
  penetrates clouds, which is important for the often-cloudy region covered
  by the dataset. Our goal for this application is to assess how well the 
  selections made by each algorithm correlate with outliers determined by
  visual inspection of the satellite images.
 \todo{If we have time, I want to also evaluate the downstream classifier
 performance before and after removing top N outliers with DORA.}

@@

\subsubsection{MNIST and FashionMNIST.}
We added this to demonstrate the capabilities of DORA with a traditional
benchmark dataset. \todo{Bryce. TBD if we will end up including due to
space and/or relevance.}

@@

\begin{table}
  \caption{Precision at $N=n_{outlier}$ for three applications; the
  best result for each data set is in bold.}
  \label{tab:faw_results}
  \centering
  \begin{tabular}{l|ccc}
    \hline
    Algorithm & Astrophysics & Earth & Planetary \\
    \hline
    PCA               & & 0.41 & 0.44 \\
    DEMUD             & & 0.41 & 0.44 \\
    \hline
    RX                & & 0.43 & ---- \\
    LRX               & & ---- & 0.33 \\
    \hline
    Isolation Forest  & & 0.46 & \textbf{0.56} \\
    \hline
    PAE               & & 0.41 & 0.22 \\
    Neg. Sampling & & \textbf{0.49} & 0.33 \\
    \hline
    Random            & & 0.32 & 0.11 \\
    \hline
  \end{tabular}
\end{table}

>>>>

\subparagraph{Probability.}
The negative sampling algorithm is implemented by converting the unsupervised 
outlier ranking problem into a semi-supervised 
problem~\citep{sipple:neg-sampling20}. Negative (anomalous) 
examples are created by sampling from an expanded space defined by the minimum 
and maximum values of each dimension of the positive (normal) examples. The 
negative and positive examples are then used to train a random forest 
classifier. We use the posterior probabilities of the random forest classifier 
as outlier scores, which means that the observations with higher posterior 
probabilities are more likely to be outliers. 
The probabilistic autoencoder is a generative model consisting of an 
autoencoder trained to reconstruct input data which is interpreted
probabilistically after training using a normalizing flow on the autoencoder
latent space~\cite{bohm2020probabilistic}. 
Samples are scored as outliers using the log likelihood in the latent
distribution, the autoencoder reconstruction error, or a combination of
the two.

@@

\subsubsection{Satellite time series for ground-truth observations.}
Many Earth science applications using satellite Earth observation (EO) data
 require ground-truth observations for understanding and modeling ground-
 identified objects in the satellite observations. These ground observations
 also serve as labels that are paired with satellite data inputs for machine 
 learning models. For example, a model trained to classify crop types in
 satellite observations requires geo-referenced labels of crop type annotated
 in the field~\cite{tseng2021learning}. A widespread challenge of using ground-
 annotated datasets is that there are often points with erroneous location or 
 label information (e.g., due to GPS location error or human entry error) that
 need to be cleaned before the dataset can be used for machine learning or
 other downstream use cases. Automatically detecting these outliers could save 
 substantial time required for cleaning datasets and improve the performance of
 downstream analyses that rely on high-quality datasets. 
 %
 We used a dataset of ground
 observations of fall army worm (FAW) in maize crops collected by the UN Food
 and Agriculture Organization (FAO). This dataset includes 6,757 samples
 with location (latitude/longitude) metadata primarily in Africa and Southeast
 Asia. Most locations coincide with crop
 fields but there are many outliers that coincide with other land cover types 
 such as water, buildings, or forests. We constructed an evaluation set 
 consisting of all samples in Kenya ($n=113$) and manually annotated whether 
 each sample was considered an inlier (i.e., located in a crop field) or not 
 using high-resolution Planet satellite images in Collect Earth 
 Online~\cite{planetdata} ($n_{inlier}=76$, $n_{outlier}=37$). We used Google
 Earth Engine to export the Sentinel-1 synthetic aperture radar (SAR) monthly
 time series for each sample location, using data from year each sample was 
 collected. We used SAR data because it is sensitive to ground texture and
  penetrates clouds, which is important for the often-cloudy region covered
  by the dataset. Our goal for this application is to assess how well the 
  selections made by each algorithm correlate with outliers determined by
  visual inspection of the satellite images.
% descoped for lack of time and space
% \todo{If we have time, I want to also evaluate the downstream classifier
% performance before and after removing top N outliers with DORA.}

@@

\subsubsection{MNIST and FashionMNIST.}
We added this to demonstrate the capabilities of DORA with a traditional
benchmark dataset. \todo{Bryce.}

@@

\begin{table}
  \caption{Precision at $N=n_{outlier}$ for three applications; the
  best result for each data set is in bold.}
  \label{tab:faw_results}
  \centering
  \begin{tabular}{l|cccc}
    \hline
    Algorithm & Astro. & Earth & Planetary & FMNIST \\
    \hline
    PCA               & & 0.41 & 0.44 \\
    DEMUD             & & 0.41 & 0.44 \\
    \hline
    RX                & & 0.43 & ---- \\
    LRX               & & ---- & 0.33 \\
    \hline
    Iso. Forest  & & 0.46 & \textbf{0.56} \\
    \hline
    PAE               & & 0.41 & 0.22 \\
    Neg. Sampling & & \textbf{0.49} & 0.33 \\
    \hline
    Random            & & 0.32 & 0.11 \\
    \hline
  \end{tabular}
\end{table}
